Compact Core Galaxies in the RESOLVE Survey

### 2016-04-18 paperdraft.tex
adding more plots

\title{Compact Core Galaxies in the RESOLVE Survey}
\author{Elaine M. Snyder\altaffilmark{1}}
\author{Sheila J. Kannappan\altaffilmark{1}}
\author{Dara J. Norman\altaffilmark{2}}
\author{Mark N. Norris\altaffilmark{3}}
\author{Callie Hood\altaffilmark{1}}
\author{Ashley S. Bittner\altaffilmark{1}}
\author{Kathleen D. Eckert\altaffilmark{1}}
\author{Amanda J. Moffett\altaffilmark{1,4}}
\author{David V. Stark\altaffilmark{1}}
\author{the RESOLVE Team}
\affil{$^1$Department of Physics and Astronomy, University of North Carolina, 141 Chapman Hall CB 3255, Chapel Hill, NC 27599,
USA}
\affil{$^2$National Optical Astronomical Observatory, 950 N Cherry Ave, Tucson, AZ 85719, USA}
\affil{$^3$Jeremiah Horrocks Institute, University of Central Lancashire, Preston, PR1 2HE, United Kingdom}
\affil{$^4$International Centre for Radio Astronomy Research (ICRAR), The University ofWestern Australia, 35 Stirling High-
way, Crawley, WA 6009, Australia}
\begin{abstract}
We analyze the complete set of 96 compact core galaxies (CCGs) in the volume-limited RESOLVE survey to investigate the formation and evolution of compact galaxies across a broad statistical distribution of environments. We perform one and two component S\'ersic fits on the galaxies in the RESOLVE sample using \textsc{Galfit} to produce seeing-deconvolved effective radius (\Reff) measurements. These one component fits are used to select candidate CCGs with \Reff\ $< 1000$ pc, and we subsequently perform new two component fits to these galaxies with the S\'ersic n of the second component held fixed at 1 for an exponential disk. We finalize the CCG sample for those galaxies with core-only (first component) radius $<800$pc, an upper limit that encompasses all of the similarly compact stellar systems (ultra compact dwarfs, cEs, nucleated dwarf ellipticals) included in the Archive of Intermediate Mass Stellar Systems (AIMSS) catalog. We quantify the amount of light in the core and envelope of each CCG by performing fixed two S\'ersic fits on UKIDSS $Y$-band imaging, and taking the ratio of the fluxes in each component. The result is a smooth continuum of CCGs that range from envelope-dominated to core+envelope to core-dominated. With GALEX $NUV$, SDSS $ugriz$, and 2MASS/UKIDSS $YJHK$ data, we derive the colors and star formation histories and find that a significant number of CCGs live on the blue sequence and have recently formed stars. We find that CCGs naturally occur in a range of environments from isolated to cluster. We compare velocity dispersions of CCGs derived from Gemini IFU data and SOAR spectroscopy to velocity dispersions of other RESOLVE galaxies. We search for CCGs offset to higher or lower dispersion in the dispersion-stellar mass relation, which may indicate tidal stripping or dissipative formation, respectively. Initial results show several CCGs following the dissipative formation track.
\end{abstract}

@@

Meanwhile, cEs have radii ranging typically from $\sim$100--600 pc (again see Figure 11 in \citet{Norris2014}). They are typically thought to be either the result of a tidal stripping event, as is the case for the prototypical cE, M32 (\citet{Choi2002}, \citet{Graham2002}), or as a part of the low mass extension to the elliptical galaxy population. The evidence for tidal stripping includes finding tidal streams near cEs (\citet{SmithCastelli2008a}, \citet{Chilingarian2009}) and one cE, studied in \citet{Kormendy1997} that is believed to host a central massive black hole. On the other hand, \citet{Kormendy2012a} argue that cEs are the extension of normal elliptical galaxies.

@@

Clues to how a CSS is formed may be given by their environments. UCDs, cEs, dEs have all been found in clusters (for UCDs:\citet{Price2009}, \citet{Madrid2010}, \citet{Hasegan2005}, \citet{Jones2006}, \citet{Mieske2009}, \citet{Wehner2007}, \citet{Misgeld2008}; for cEs: \citet{Chilingarian2007}, \citet{SmithCastelli2012}, \citet{Price2009}; and dEs: \citep{SmithCastelli2012}, \citet{Koo1994}, \citet{Guzman1996}, \citet{Crawford2016}). Fewer CSSs have been found in groups: \citet{Evstigneeva2007} find one definite UCD and two candidates in the Dorado group as well as two candidates in the NGC1400 group and \citet{Hau2009} find a UCD near the Sombrero galaxy; \citet{Huxor2011} and \citet{Chilingarian2010} find cEs in groups; and \citet{Crnojevi2014}  \citet{Penny2014} find dEs in groups. Lastly, only cEs have been found in isolated environments: \citet{Huxor2013}, \citet{Paudel2014}, and \citet{Chilingarian2015} find cEs separated from other galaxies by $>1$ Mpc.

Given the similarities in formation scenarios and environments for these CSSs, we propose to study these galaxies as a part of a continuum of compact galaxies instead of being broken up into (perhaps somewhat arbitrary) classes based on \Reff\ sizes. Critical to this study is a ubiquitous sample of CSSs that will allow us to analyze their formation as a function of all environments. Previous studies have mainly focused on selecting targets in the group/cluster environments, since this route is often more convenient. Our goal is thus to create a complete and volume-limited sample of compact galaxies across all environments so that we may examine their formation and evolution.

@@

\noindent We use the Archive of Intermediate Mass Stellar Systems (AIMSS) catalog (\citet{Norris2014}, \citet{Forbes2014}, \citet{Janz2015}) as a reference sample for our CCGs.  AIMSS is a catalog of spectroscopically confirmed CSSs found in groups and clusters and includes both literature and newly discovered objects. Their process to find new CSSs is as follows. First, a search is conducted in the Hyperleda redshift catalog \citep{Paturel2003} for all galaxies at distances between $\sim 7$ and 200 Mpc (to ensure CSSs with \Reff\ $> 50$ pc will be resolved in any available HST imaging) and with M$_{\rm B} < -15$. Once complete, the team then searches the Hubble Space Telescope archive for WFCP2, ACS, and WFC3 imaging within 150kpc in projection of the selected galaxies. Next, Source Extractor is run on the HST images, and candidates are identified using a training set of previously known CSSs from the literature. Once cross-matched to ensure none are previously known and vetted by eye, spectra are obtained, redshift confirmation is performed, and velocity dispersions are extracted. AIMSS also includes literature data for (nucleated) dwarf ellipticals, dwarf spheroidals, dwarf S0s, young massive clusters, and massive elliptical galaxies. AIMSS compiles M$_{\rm V}$, stellar mass, effective radius, and velocity dispersions for each of its objects. Thus, though not a statistically-defined sample, AIMSS provides a useful reference catalog to which we can calibrate and compare our CCGs.

@@

\noindent We begin our analysis by deriving seeing-corrected \Reff\ measurements for all galaxies in the RESOLVE sample. For this task we employ \textsc{Galfit} \citep{Peng2002} to deconvolve the seeing point spread function (PSF) and to perform one and two component fits on each galaxy. Deconvolution is especially important since the smallest galaxies in RESOLVE are most effected by seeing, and accurate radii measurements are crucial for the study of these objects.

@@

\item[The initial input file]{The initial input file tells \textsc{Galfit} where to find the image, PSF, mask, and constraint, and also holds the initial guesses for our fit parameters. These initial parameters can just be rough guesses, but for our purposes, we use the previously derived photometric data for RESOLVE galaxies as our initial guesses (see \autoref{phot} for more details). For the one component fits, we set up our input file to have a section for measuring the level of the sky background, and another section for fitting a S\'ersic profile to galaxy. The S\'ersic model will find a best fit for the following parameters: central x \& y position of the model, integrated magnitude, \Reff, S\'ersic n, axial ratio (b/a), and position angle.}

\item[Imaging data]{We use publicly-available high resolution $Y$-band imaging from the UKIDSS Large Area Survey \citep{Lawrence2007}. This data set was chosen because it covers both RESOLVE-A and RESOLVE-B and offers $\sim0.8''$ seeing resolution corresponding to $\sim350$pc at RESOLVE distances in turn letting us span down to the sizes of cE-like galaxies. The $Y$ band is advantageous too, since it probes the underlying stellar population and will not skew our measurements to larger radii if the galaxy is currently experiencing a starburst. We download $\sim10'\times10'$ images of each galaxy from the WFCAM Science Archive\footnote{\url{was.roe.ac.uk//}}}. While \textsc{Galfit} does not require images this large to operate normally, we choose this size so that there will be more stars in the field of view which is useful for constructing a PSF. There are  only XX galaxies in RESOLVE that do not have UKIDSS imaging and thus are excluded from our analysis. Visual inspection of these galaxies indicate that they would most likely not be classified as CCGs anyways.

\item[The PSF image]{We build our PSF image via the following steps.
\begin{enumerate}

\item We begin by running Source Extractor on the image to identify every object in the frame, including stars, galaxies, and even sometimes hot pixels. Source Extractor outputs a .txt file that has estimates of each object's flux, elongation, ellipticity, radius, and \texttt{class\_star}.

\item With this information, we select stars to use for building our PSF. The limiting parameters are (1) that the flux must be between $5,000-100,000$ counts to ensure no extremely faint or saturated stars are included, (2) that the ellipticity be lt 0.2 to ensure we are selecting round objects, (3) that the effective radius of the star be greater than 1 pixel to avoid selecting hot pixels, and (4) that \texttt{class\_star} is greater than 0.6 to ensure the objects we're picking are most likely stars. We also make sure the stars are not near the image edges, since there can sometimes be image artifacts there. Lastly, we make sure to not include the galaxy we are going to fit, since the cE-like galaxies can appear to be stars.

@@

\item We also use the background median to reject any of the cutouts that have a background median greater than $2\sigma$ from the mean of all the background median values. This ensures we are not including any cutouts that have stars with large or bright neighboring objects. We are now left with background-subtracted cutouts of stars with no neighbors that could effect our stacking.

\item Before we can combine all the cutouts, we must normalize the flux of each central star to the same amplitude. To do this, we run each cutout through Source Extractor once more to obtain the fluxes of each central star, and use these to weight the cutouts accordingly. We calculate the weight for each cutout as follows: (1) we find the central star with the highest flux of all the cutouts and then (2) divide every flux by this maximum flux. We then divide each cutout by its corresponding weight, which makes each central star have the same amplitude.

@@

\noindent Our last step is to fit a 2D Gaussian to the PSF to determine a rough estimate seeing full width half max (FWHM), as a sanity check. We find that the average FWHM for all PSFs is $\sim 0.9''$, only slightly worse than the $\sim 0.8''$ seeing expected for UKIDSS imaging. \textsc{Galfit} uses this PSF image to deconvolve the galaxy light profile from the seeing, thereby removing the effects of atmospheric scattering of the galaxy light as the data was taken.}

\item[The mask image]{The first step in making the mask is to create an image that has the same dimensions as our input image but with each pixel value equaling zero. We again use Source Extractor to compile a list of all the objects in the input image, and  use this list to replace the pixel values at the coordinates of the located objects with one. \textsc{Galfit} uses the mask image to ignore any objects neighboring our galaxy while it is completing the fit.}

\item[The constant file]{The constraint file hosts the limits we are willing to accept for each of the fit parameters. The constraints for x \& y position in the image, integrated magnitude, \Reff, S\'ersic n, axial ratio (b/a), and position angle for the one component fits are listed in \autoref{constrtable}. }

@@

\noindent We use all of the above to obtain a one component, deconvolved fit for every galaxy in the RESOLVE survey. We then perform the two component fits as follows. We use the output of the one component fit as the basis of the input file for the two component fit. The first two sections in the two component fit initial input file will be identical to the output of the one component fit (except that the names of the files to be output are changed). We then add a third section to include a second S\'ersic fit, and change the initial guess for \Reff\ to be double that of the one component best fit. This gives \textsc{Galfit} an idea to fit on both a core and envelope. The PSF, image, and mask image will all be the same as before. We add to the constraint file the same constraints as listed in \autoref{constrtable}, but for the third section. In other words, there will be two sets of constraints in the file now, with the first set applying to the first S\'ersic fit and the second set applying the second S\'ersic fit. The constraints for each parameter are the same for both sets. We do not further constrain the two component fits since we are looking for general fits to the wide variety of galaxies in RESOLVE and do not want to over constrain our fits.

@@

\subsection{Kinematics}
\label{kin}

>>>>

\title{Compact Core Galaxies in the RESOLVE Survey}
\author{Elaine M. Snyder\altaffilmark{1}}
\author{Sheila J. Kannappan\altaffilmark{1}}
\author{Dara J. Norman\altaffilmark{2}}
\author{Mark N. Norris\altaffilmark{3}}
\author{Callie Hood\altaffilmark{1}}
\author{Ashley S. Bittner\altaffilmark{1}}
\author{Kathleen D. Eckert\altaffilmark{1}}
\author{Amanda J. Moffett\altaffilmark{1,4}}
\author{David V. Stark\altaffilmark{1}}
\author{the RESOLVE Team}
\affil{$^1$Department of Physics and Astronomy, University of North Carolina, 141 Chapman Hall CB 3255, Chapel Hill, NC 27599, USA; emsnyder@live.unc.edu}
\affil{$^2$National Optical Astronomical Observatory, 950 N Cherry Ave, Tucson, AZ 85719, USA}
\affil{$^3$Jeremiah Horrocks Institute, University of Central Lancashire, Preston, PR1 2HE, United Kingdom}
\affil{$^4$International Centre for Radio Astronomy Research (ICRAR), The University ofWestern Australia, 35 Stirling High-
way, Crawley, WA 6009, Australia}
\begin{abstract}
We analyze the complete set of 96 compact core galaxies (CCGs) in the volume-limited RESOLVE survey to investigate the formation and evolution of compact galaxies such as ultra compact dwarfs (UCDs), compact ellipticals (cEs), and dwarf ellipticals (dEs) across a broad statistical distribution of environments. We perform one and two component S\'ersic fits KIDSS for all galaxies in the RESOLVE sample with UKIDSS $Y$-band imaging using \textsc{Galfit}, which produces seeing-deconvolved effective radius (\Reff) measurements. The one component fits are used to select candidate CCGs with \Reff\ $< 1000$ pc, and then perform new two component fits to these galaxies with the S\'ersic n of the second component held fixed at 1 for an exponential disk. We finalize the CCG sample by selecting on core-only (first component) \Reff\ $<800$pc, an upper limit that encompasses all of the similarly compact stellar systems (UCDs, cEs, dEs) included in the Archive of Intermediate Mass Stellar Systems (AIMSS) catalog. We quantify the amount of light in the core and envelope of each CCG by taking the logarithm of the ratio of the fluxes in each component. The result is a smooth continuum of CCGs that range from envelope-dominated to core+envelope to core-dominated. With GALEX $NUV$, SDSS $ugriz$, and 2MASS/UKIDSS $YJHK$ data, we derive colors and star formation histories and find that a significant number of CCGs live on the blue sequence and have recently formed stars. We find that CCGs naturally occur in a range of environments from isolated to cluster. We compare velocity dispersions of CCGs derived from Gemini IFU data and SOAR spectroscopy to velocity dispersions of other RESOLVE galaxies and the cEs and and dEs of AIMSS. We search for CCGs offset to higher or lower dispersion in the dispersion-stellar mass relation, which may indicate tidal stripping or dissipative formation, respectively. Initial results show several CCGs following the dissipative formation track.
\end{abstract}

@@

Meanwhile, cEs have radii ranging typically from $\sim$100--600 pc (again see Figure 11 in \citet{Norris2014}). They are typically thought to be either the result of a tidal stripping event, as is the case for the prototypical cE, M32 (\citet{Choi2002}, \citet{Graham2002}), or as a part of the low mass extension to the E population. The evidence for tidal stripping includes finding tidal streams near cEs (\citet{SmithCastelli2008a}, \citet{Chilingarian2009}) and one cE, studied in \citet{Kormendy1997}, that is believed to host a central massive black hole. On the other hand, \citet{Kormendy2012a} argue that cEs are the extension of normal E galaxies.

@@

Clues to how a CSS is formed may be given by their environments. UCDs, cEs, dEs have all been found in clusters (for UCDs: \citet{Price2009}, \citet{Madrid2010}, \citet{Hasegan2005}, \citet{Jones2006}, \citet{Mieske2009}, \citet{Wehner2007}, \citet{Misgeld2008}; for cEs: \citet{Chilingarian2007}, \citet{SmithCastelli2012}, \citet{Price2009}; and dEs: \citep{SmithCastelli2012}, \citet{Koo1994}, \citet{Guzman1996}, \citet{Crawford2016}). Fewer CSSs have been found in groups: \citet{Evstigneeva2007} find one definite UCD and two candidates in the Dorado group as well as two candidates in the NGC1400 group and \citet{Hau2009} find a UCD near the Sombrero galaxy; \citet{Huxor2011} and \citet{Chilingarian2010} find cEs in groups; and \citet{Crnojevi2014}  \citet{Penny2014} find dEs in groups. Lastly, only cEs have been found in isolated environments: \citet{Huxor2013}, \citet{Paudel2014}, and \citet{Chilingarian2015} find cEs separated from other galaxies by $>1$ Mpc.

Given the similarities in formation scenarios and environments for these CSSs, we propose to study these galaxies as a part of a continuum of compact galaxies instead of being broken up into (perhaps somewhat arbitrary) classes based on \Reff. Critical to this study is a ubiquitous sample of CSSs that will allow us to analyze their formation as a function of all environments. Previous studies have mainly focused on selecting targets in the group/cluster environments, since this route is often more convenient. Our goal is thus to create a complete and volume-limited sample of compact galaxies across all environments so that we may examine their formation and evolution.

@@

\noindent We use the Archive of Intermediate Mass Stellar Systems (AIMSS) catalog (\citet{Norris2014}, \citet{Forbes2014}, \citet{Janz2015}) as a reference sample for our CCGs.  AIMSS is a catalog of spectroscopically confirmed CSSs found in groups and clusters and includes both literature and newly discovered objects. Their process to find new CSSs is as follows. First, a search is conducted in the Hyperleda redshift catalog \citep{Paturel2003} for all galaxies at distances between $\sim 7$ and 200 Mpc (to ensure CSSs with \Reff\ $> 50$ pc will be resolved in any available HST imaging) and with M$_{\rm B} < -15$. Once complete, the team then searches the Hubble Space Telescope archive for WFCP2, ACS, and WFC3 imaging within 150kpc in projection of the selected galaxies. Next, Source Extractor is run on the HST images, and candidates are identified using a training set of previously known CSSs from the literature. Once cross-matched to ensure none are previously known and vetted by eye, spectra are obtained, redshift confirmation is performed, and velocity dispersions are extracted. AIMSS also includes literature data for other UCDs, cEs, (nucleated) dEs, dwarf spheroidals, dwarf S0s, young massive clusters, and Es. AIMSS compiles M$_{\rm V}$, stellar mass, effective radius, and velocity dispersions for each of its objects. Thus, though not a statistically-defined sample, AIMSS provides a useful reference catalog to which we can calibrate and compare our CCGs.

@@

\noindent We begin our analysis by deriving seeing-deconvolved \Reff\ measurements for all galaxies in the RESOLVE sample. For this task we employ \textsc{Galfit} \citep{Peng2002} to deconvolve the seeing point spread function (PSF) and to perform one and two component fits on each galaxy. Deconvolution is especially important since the smallest galaxies in RESOLVE are most effected by seeing, and accurate radii measurements are crucial for the study of these objects.

@@

\item[The initial input file]{The initial input file gives \textsc{Galfit} the files paths to the image, PSF, mask, and constraint file, and also holds the initial guesses for our fit parameters. For the one component fits, we set up our input file to have a section for measuring the level of the sky background, and another section for fitting a S\'ersic profile to galaxy. The S\'ersic model will find a best fit for the following parameters: central x \& y position of the model, integrated magnitude, \Reff, S\'ersic n, axial ratio (b/a), and position angle. These initial values for these parameters can just be rough guesses, but for our purposes, we use values for the magnitude, \Reff, b/a, and PA that are derived from RESOLVE's photometric data (see \autoref{phot} for more details)}

\item[Imaging data]{We use publicly-available high resolution $Y$-band imaging from the UKIDSS Large Area Survey \citep{Lawrence2007}. This data set was chosen because it covers both RESOLVE-A and RESOLVE-B and offers $\sim0.8''$ seeing resolution corresponding to $\sim350$pc at RESOLVE distances in turn letting us span down to the sizes of cE-like galaxies. The $Y$ band is advantageous too, since it probes the underlying stellar population and will not skew our measurements to larger radii if the galaxy is currently experiencing a starburst. We download $\sim10'\times10'$ images of each galaxy from the WFCAM Science Archive\footnote{\url{was.roe.ac.uk//}}}. While \textsc{Galfit} does not require images this large to operate normally, we choose this size so that there will be more stars in the field of view which is useful for constructing a PSF. There are only 24 galaxies in RESOLVE that do not have UKIDSS imaging and thus are excluded from our analysis.

\item[The PSF image]{\textsc{Galfit} uses the PSF image to deconvolve the galaxy light profile from the seeing, thereby removing the effects of atmospheric scattering of the galaxy light as the data was taken. We build our PSF image via the following steps.
\begin{enumerate}

\item We begin by running Source Extractor on the image to identify every object in the frame, including stars, galaxies, and even sometimes hot pixels. Source Extractor outputs a text file that has estimates of each object's flux, elongation, ellipticity, radius, and \texttt{class\_star}.

\item With this information, we select stars to use for building our PSF. The limiting parameters are (1) that the flux must be between $5,000-100,000$ counts to ensure no extremely faint or saturated stars are included, (2) that the ellipticity be lt 0.2 to ensure we are selecting round objects, (3) that the effective radius of the star be greater than 1 pixel to avoid selecting hot pixels, and (4) that \texttt{class\_star} is greater than 0.6 to ensure the objects we're picking are most likely stars. We also make sure the stars are not near the image edges, since there can sometimes be image artifacts there. Lastly, we make sure to not include the galaxy we are going to fit, since the cE-like galaxies can sometimes appear to be stars.

@@

\item We then reject any of the cutouts that have a background median value greater than $2\sigma$ from the mean of the background median values for every cutout. This ensures we are not including any cutouts that have stars with large or bright neighboring objects. We are now left with background-subtracted cutouts of stars with no neighbors that could affect our stacking.

\item Before we can combine all the cutouts, we must normalize the flux of each central star to the same amplitude. To do this, we run each cutout through Source Extractor once more to obtain the fluxes of each central star and use these to weight the cutouts accordingly. We calculate the weight for each cutout as follows: (1) we find the central star with the highest flux of all the cutouts and then (2) divide every flux by this maximum flux. We then divide each cutout by its corresponding weight, which makes each central star have the same amplitude.

@@

\noindent Our last step is to fit a 2D Gaussian to the PSF to determine a rough estimate seeing full width half max (FWHM), as a sanity check. We find that the average FWHM for all PSFs is $\sim 0.9''$, only slightly worse than the $\sim 0.8''$ seeing expected for UKIDSS imaging.}

\item[The mask image]{\textsc{Galfit} uses the mask image to ignore any objects neighboring our galaxy while it is completing the fit.The first step in making the mask is to create an image that has the same dimensions as our input image but with each pixel value equaling zero. We again use Source Extractor to compile a list of all the objects in the input image, and  use this list to replace the pixel values at the coordinates of the located objects with one.}

\item[The constant file]{The constraint file hosts the limits we are willing to accept for each of the fit parameters. The constraints for x \& y position in the image, integrated magnitude, \Reff, S\'ersic n, axial ratio (b/a), and position angle for the one component fits are listed in \autoref{constrtable}. We change the constraint file for the two component fits later in the analysis.}

@@

\noindent We use all of the above to obtain a one component, seeing-deconvolved fit for every galaxy in the RESOLVE survey. We then perform the two component fits as follows. We use the output of the one component fit as the basis of the input file for the two component fit. The first two sections in the two component fit initial input file will be identical to the output of the one component fit (except that the names of the files to be output are changed). We then add a third section to include a second S\'ersic fit, and change the initial guess for \Reff\ to be double that of the one component best fit. This gives \textsc{Galfit} an idea to fit on both a core and envelope. The PSF, image, and mask image will all be the same as before. We add to the constraint file the same constraints as listed in \autoref{constrtable}, but for the third section. In other words, there will be two sets of constraints in the file now, with the first set applying to the first S\'ersic fit and the second set applying the second S\'ersic fit. The constraints for each parameter are the same for both sets. We do not further constrain the two component fits since we are looking for general fits to the wide variety of galaxies in RESOLVE and do not want to over constrain our fits.

@@

\subsection{Kinematics}
\label{kindata}
